\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2011-05-06 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
added incomplete dissociation and spectral counting sections to demo vignette and mention bio-sig-proteomics list
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@55300 bc3139a8-67e5-0310-9ffc-ced21a209358

<<MSnExp,cache=FALSE,echo=TRUE>>=
library("MSnbase")
filename <- dir(system.file(package="MSnbase",dir="extdata"),
                full.name=TRUE,pattern="mzXML$")
print(filename)
experiment <- readMzXMLData(filename, msLevel = 2, verbose=FALSE)
experiment
@

<<experiment-size,echo=FALSE,cache=FALSE>>=
sz <- sum(sapply(assayData(experiment),object.size)) + object.size(experiment)
sz <- round(sz/(1024^2),2)
@

@@

The raw data is composed of the \Sexpr{length(experiment)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(experiment)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{experiment[["X10"]]} or as a list with \texttt{spectra(experiment)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

<<Spectrum,cache=FALSE,echo=TRUE>>=
sp <- experiment[["X43"]]
sp
@

@@

<<accessors,cache=FALSE,echo=TRUE>>=
peaksCount(sp)
head(peaksCount(experiment))
rtime(sp)
head(rtime(experiment))
@

@@

<<extractPrecSpec,echo=T,cache=F>>=
exp2 <- extractPrecSpectra(experiment,precursorMz(sp))
exp2
@

@@

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(experiment,t=3,verbose=FALSE)
## total ion current
tic(sp)
tic(experiment[["X43"]])
@

@@

Quantitation is performed on specific peaks in the spectra, that are specified 
with an \Robject{ReporterIon} object. A specific peak is defined by it's expected 
\Robject{mz} value and is expected to be found within \Robject{mz} $\pm$ \Robject{width}.  
If no data is found, \Robject{NA} is returned.

@@

\begin{figure}[!ht]
<<cv-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
plot(order(mn1),cv1,col="steelblue",pch=19,
     ylab="CV",xlab="Rank mean")
points(order(mn2),cv2,col="red",pch=19)
grid()
lines(seq(1,70,by=(70/length(rmed1))),rmed1,col="steelblue",lwd=2)
lines(seq(1,70,by=(70/length(rmed2))),rmed2,col="red"      ,lwd=2)
## legend("topright",
##        legend=c("log2","vsn","running\nmedian"),
##        pch=c(19,19,NA),
##        col=c("steelblue","red","black"),
##        lwd=c(0,0,2),
##        bty="n",
##        cex=.7)
@ 
\caption{CV versus signal intensity comparison for log2 (blue) and vsn (red) transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}

\subsection{Feature aggregation}

@@

>>>>

<<MSnExp,cache=FALSE,echo=TRUE>>=
library("MSnbase")
filename <- dir(system.file(package="MSnbase",dir="extdata"),
                full.name=TRUE,pattern="mzXML$")
print(filename)
raw.experiment <- readMzXMLData(filename, msLevel = 2, verbose=FALSE)
raw.experiment
@

<<experiment-size,echo=FALSE,cache=FALSE>>=
sz <- sum(sapply(assayData(raw.experiment),object.size)) + 
  object.size(raw.experiment)
sz <- round(sz/(1024^2),2)
@

@@

The raw data is composed of the \Sexpr{length(raw.experiment)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(raw.experiment)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{raw.experiment[["X10"]]} or as a list with \texttt{spectra(raw.experiment)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

<<Spectrum,cache=FALSE,echo=TRUE>>=
sp <- raw.experiment[["X43"]]
sp
@

@@

<<accessors,cache=FALSE,echo=TRUE>>=
peaksCount(sp)
head(peaksCount(raw.experiment))
rtime(sp)
head(rtime(raw.experiment))
@

@@

<<extractPrecSpec,echo=T,cache=F>>=
exp2 <- extractPrecSpectra(raw.experiment,precursorMz(sp))
exp2
@

@@

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(raw.experiment,t=3,verbose=FALSE)
## total ion current
tic(sp)
tic(experiment[["X43"]])
@

@@

Quantitation is performed on fixed peaks in the spectra, that are specified 
with an \Robject{ReporterIon} object. A specific peak is defined by it's expected 
\Robject{mz} value and is searched for within \Robject{mz} $\pm$ \Robject{width}.  
If no data is found, \Robject{NA} is returned.

@@

\begin{figure}[!ht]
<<cv-plot,fig=T,echo=F,tikz=T,external=F,width=5,height=5>>=
plot(order(mn1),cv1,col="steelblue",pch=19,
     ylab="CV",xlab="Rank mean",
     cex.axis=.7,cex.lab=.8)
points(order(mn2),cv2,col="red",pch=19)
grid()
lines(seq(1,70,by=(70/length(rmed1))),rmed1,col="steelblue",lwd=1)
lines(seq(1,70,by=(70/length(rmed2))),rmed2,col="red"      ,lwd=1)
## legend("topright",
##        legend=c("log2","vsn","running\nmedian"),
##        pch=c(19,19,NA),
##        col=c("steelblue","red","black"),
##        lwd=c(0,0,2),
##        bty="n",
##        cex=.7)
@ 
\caption{CV versus signal intensity comparison for log2 (blue) and vsn (red) transformed data. Lines indicate running CV medians.}
\label{fig:cv-plot}
\end{figure}

\subsection{Feature aggregation}
\label{sec:feataggregation}

@@

\subsection{Label-free MS2 quantitation}
\label{sec:spectralcounting}

Note that if samples are not multiplexed, label-free MS2 quantitation is possible 
using \Rpackage{MSnbase}.
Once individual spectra have been assigned to peptides and proteins, it becomes 
straightforward to estimate protein quantities using the spectral counting method, 
as illustrated in section \ref{sec:feataggregation}, when the \Robject{groupBy} 
argument is defined.

\section{Quantitative assessment of incomplete dissociation}
\label{sec:incompdissoc}

Quantitation using isobaric reporter tags assumes complete dissociation between
the reporter group (red on figure \ref{fig:itraqchem}), 
balance group (blue) 
and peptide (the peptide reactive group is drawn in green). 
However, incomplete dissociation does occur and 
results in an isobaric tag (i.e reporter and balance groups) specific peaks.

\begin{figure}[h]
  \begin{center}
    \includegraphics[width=4cm]{./itraqchem.pdf}
\caption{ iTRAQ 4-plex isobaric tags reagent consist of three parts: 
  (1) a charged reporter group (MZ of 114, 115, 116 and 117) that is unique to each of the four reagents (red),
  (2) an uncharged mass balance group (28-31 Da) (blue)and 
  (3) a peptide reactive group (NHS ester) that binds to the peptide. 
  In case of incomplete dissociation, the reporter and balance groups produce a specific peaks 
  at MZ 145.
}
\label{fig:itraqchem}
  \end{center}
\end{figure}

\Rpackage{MSnbase} provides, among others, a \Robject{ReporterIons} object for iTRAQ 4-plex that includes the 
145 peaks, called \Robject{iTRAQ5}. This can then be used to quantify the experiment as show in section \ref{sec:quant} 
to estimate incomplete dissociation for each spectrum.

<<incompdiss,echo=TRUE,cache=FALSE,keep.source=T>>=
iTRAQ5
incompdiss <- quantify(raw.experiment,
                       method="trap",
                       reporters=iTRAQ5,
                       strict=FALSE,
                       verbose=FALSE)
head(exprs(incompdiss))
@

Figure \ref{fig:incompdiss} compares these intensities for the whole experiment.

\begin{figure}[!hbt]
  \begin{center}
<<incompdiss-plot,fig=T,echo=F,tikz=T,external=F,width=4,height=3.5>>=
dfr <- melt(exprs(incompdiss))
colnames(dfr) <- c("feature","reporters","intensity")
qplot(x=reporters,y=intensity,data=dfr,geom="boxplot") + theme_gray(8)
@ 
\caption{ Boxplot comparing intensities of the 4 reporter ions an the 
  incomplete dissociation specific peak. }
\label{fig:incompdiss}
  \end{center}
\end{figure}

\clearpage
